Natural selection in the Health and Retirement Study

### 2023-09-21 hrs-selection.qmd
Add conclusion para; mention regs controlling for age at first birth

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefb = r10agey_e - h10ageokid,
    agefb = ifelse(agefb < 12, NA_real_, agefb),
    agefb = chop_equally(agefb, 2, labels = c("Low", "High"))
  )

@@

rand <- rand |> 
  select(ethnicity, all_of(pgs), all_of(pcs), weight, raevbrn, rlrs,
         raedyrs, r10sibs, edu, income_med, married, agefb, rabyear,
         raehsamp, raestrat, parent_weight)

@@

1.  There is a trade-off between time spent working and raising children. This leads people with more human capital and higher expected wages to have fewer children.
2.  The trade-off is sharper for low-income people, people with low human capital, and single parents. As a result, natural selection is stronger among these groups.

The evidence for point 1 is that polygenic scores which correlate positively with human capital correlate negatively with number of children, i.e. they are being selected against. And vice versa, scores which correlate positively with human capital are being selected for. The evidence for point 2 is that correlations with number of children are larger and more significant among people with lower income, less education and single parents.

Here, I make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so I can check whether the human capital theory predicts patterns of natural selection in both these ethnicities.

@@

coefs_fert_edu <- cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
cor_fert_edu_phen <- list()

@@

  cor_fert_edu_phen[[eth]] <- 
    svyglm_quiet(rlrs ~ raedyrs, r_eth) |> 
    coef() |> 
    pluck("raedyrs")
}

@@

Appendix @fig-x-agefb splits respondents by age at first birth (median within each ethnicity). There are no large differences between younger and older parents. As in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (`r cor_agefb`), and for `r n_discordant` out of `r length(pgs)` PGS, coefficients have opposite signs. However, in bootstraps, these statistics are not significantly different from 0 and `r length(pgs)/2` respectively. Note that the N is lower here due to missing values.

@@

>>>>

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

@@

rand <- rand |> 
  select(ethnicity, all_of(pgs), all_of(pcs), weight, raevbrn, rlrs,
         raedyrs, r10sibs, edu, income_med, married, agefb, agefbn,
         rabyear, raehsamp, raestrat, parent_weight, ragender)

@@

1.  There is a trade-off between time spent working and raising children. This leads people with more human capital and higher expected wages to have fewer children. Evidence for this is that polygenic scores which correlate positively with human capital correlate negatively with number of children, i.e. they are being selected against; conversely, scores which correlate positively with human capital are being selected for.
2.  The trade-off is sharper for low-income people, people with low human capital, and single parents. As a result, natural selection is stronger among these groups. Evidence for this is that scores' regression coefficients on number of children are larger among people with lower income or less education, and single parents.

Here, I make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so I can check whether the theory predicts patterns of natural selection in both these ethnicities.

@@

coefs_fert_edu <- cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
tidy_fert_edu_phen <- list()

@@

  tidy_fert_edu_phen[[eth]] <- 
    svyglm_quiet(rlrs ~ raedyrs, r_eth) |> 
    broom::tidy(conf.int = TRUE) |> 
    filter(term == "raedyrs")
}

@@

Appendix @fig-x-agefb splits respondents by age at first birth (median within each gender). Note that the N is lower here due to missing values. There are no large differences between younger and older parents. As in the UK Biobank sample, correlations of PGS coefficients between younger and older parents are negative (`r cor_agefb`), and for `r n_discordant` out of `r length(pgs)` PGS, coefficients have opposite signs. However, in bootstraps, these statistics are not significantly different from 0 and `r length(pgs)/2` respectively. Also, in the UK Biobank sample, PGS coefficients on fertility controlling for age at first birth were negatively correlated with uncontrolled coefficient; here, that isn't true, with coefficients remaining positively correlated.

```{r}
#| label: stat-control-agefb
#| eval: false

# cselection effects ontrolling for age at first birth
# 
form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn", pcs), 
                                                    response = "rlrs"))
mod_pgs_control_agefb <- map(form_pgs_control_agefb,
                             \(x) svyglm_quiet(x, rand, subset = 
                                                 ethnicity == "white"))
tidy_pgs_control_agefb <- map(mod_pgs_control_agefb, broom::tidy) |> 
                          list_rbind() |> 
                          filter(term %in% pgs)

cor(tidy_pgs$white$estimate, tidy_pgs_control_agefb$estimate)

@@

Why does the US data show fewer differences by socio-economic status (SES) than the UK? One possibility is that SES maps on to race in the US, so that ethnic differences here capture some of the variation seen in the UK. The regression coefficient of phenotypic educational attainment on fertility is more negative among black than white respondents (blacks: `r tidy_fert_edu_phen$black$estimate`, 95% C.I. `r tidy_fert_edu_phen$black$conf.low` to `r tidy_fert_edu_phen$black$conf.high`; whites: `r tidy_fert_edu_phen$white$estimate`, 95% C.I. `r tidy_fert_edu_phen$white$conf.low` to `r tidy_fert_edu_phen$white$conf.high`; cf. @goldscheider1969minority, @johnson1979minority, @yang2003big). And the slope of PGS education coefficients on fertility coefficients is larger among black respondents, though imprecisely estimated (see @fig-rlrs-edu). But comparisons of PGS selection coefficients between the ethnic groups are hard because of the smaller sample size and by their different predictive power, so this hypothesis can only be speculative.
